Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-01-24 paper_comparison/paper_comparison/paper_comparison.Rmd
small changes

```{r}
# Get bootstrap CI
get_boot_conf <- function(dat, 
                          .x,
                          .y,
                          times = 100, 
                          upper_q = 0.975, 
                          lower_q = 0.025) {

  # Construct the model formula
  mod_formula <- substitute(.y ~ .x)

  # Get the x variable name
  x_chr <- as.character(substitute(.x))

  # Prepare a sequence of x values
  x_seq <- seq(min(dat[[x_chr]]), max(dat[[x_chr]]), by = 0.01)

  # Properly construct the new data frame for prediction
  newdata <- data.frame(.x = x_seq)
  newdata[[x_chr]] <- x_seq

  # Get the prediction for each bootstrap sample   
  dat_pred <- map_df(1:times, function(i) {

    # Get a bootstrap sample
    dat_i <- slice_sample(dat, n = nrow(dat), replace = TRUE) %>%
      mutate(boot_id = i)

    # Fit the model on the sample
    mod_i <- glm(mod_formula, family = binomial, data = dat_i)

    # Get the prediction
    newdata$.pred <- predict(mod_i, newdata = newdata, type = "response")

    # Return the prediction
    newdata
  })

  dat_pred <- dat_pred %>%
    group_by(.x) %>%
    summarise(upper = quantile(.pred, upper_q), 
              lower = quantile(.pred, lower_q))
  dat_pred[[x_chr]] <- dat_pred$.x
  dat_pred
}
```

@@

  # Draw bootstrap CI for visual test
  geom_ribbon(data = ~get_boot_conf(.x,
                                    log_effect_size,
                                    reject,
                                    times = 500),
              aes(x = log_effect_size, 
                  ymin = lower, 
                  ymax = upper),
              fill = rev(rcartocolor::carto_pal(4, "Vivid"))[4],
              alpha = 0.2) +

@@

```{r polypower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for non-linear patterns (uniform fitted values only). Main plot shows the power curves estimated using logistic regression, with dots indicating human evaluations of lineups. Surrounding band of the visual test shows the 95% confidence interval estimated using 500 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much non-linearity in the residuals? For the RESET test this is around log effect size 1.5, but for the visual test it is around 3.5."}

@@

```{r heterpower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating human evaluations of lineups. Surrounding band of the visual test shows the 95% confidence interval estimated using 500 bootstrap samples. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much heteroskedasticity in the residuals? For the BP test this is around log effect size 1.5, but for the visual test it is around 3."}

@@

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. Although the power of visual test under the discrete uniform distribution is the lowest in the case of non-linearity patterns, it has a relatively great power this time. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Visual tests have low power on lineups with the normal distribution and the lognormal distribution. This suggests the use of transformation of fitted values in reading residual plots. In the case of conventional tests, the discrete uniform distribution consistently perform well, followed by the uniform distribution, the normal distribution and the lognormal distribution.

@@

```{r different-x-dist-poly-power, fig.cap = "Power of different tests on lineups with different fitted value distributions. Each power curve of the uniform distribution is surrounded by a grey line to indicate it is the baseline within its group. Conventional tests are more sensitive than visual tests in regards of the type of the visual pattern and the fitted value distribution. Visual tests exhibit relatively low power in testing non-linearity patterns under discrete uniform distribution, possibly due to the difficulty in reading discreteness. Meanwhile, uneveness affects heteroskedasticity patterns greatly, result in low power of visual tests under normal and lognormal distribution."}

@@

```{r poly-power-j, fig.height = 4, fig.cap = "Power of conventional tests and visual tests on lineups containing four different non-linearity shapes. "}

@@

We also investigate the impact of the shape of heteroskedasticity on the power of different tests. The corresponding result is presented in Figure \ref{fig:heter-power-a}. In theory, the "Left-triangle" and the "Right-triangle" should have no difference from the point of view of a Breusch-Pagan test. According to Figure \ref{fig:heter-power-a}, this is indeed the case. And these two shapes should have the same visual test power, unless the orientation of the visual pattern is a non-negligible factor from human perspective. However, the same graph suggests that the "Right-triangle" shape is more difficult to be noticed by subjects than the "Left-triangle" shape.

@@

heter_conv_sim %>%
  filter(x_dist == "uniform") %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(reject = as.numeric(BP_p_value < 0.05)) %>%
  mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>%
  mutate(conv_or_not = "Conventional") %>%
  bind_rows(vi_survey %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  filter(x_dist == "uniform", type == "heteroskedasticity") %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(reject = as.numeric(p_value < 0.05)) %>%
  mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>%
  mutate(conv_or_not = "Visual")) %>%
  mutate(a = factor(a, levels = c("Left-triangle", "Butterfly", "Right-triangle"))) %>%
  select(a, log_effect_size, reject, conv_or_not) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject, col = a), 
              se = FALSE, 
              method = "glm",
              method.args = list(family = binomial())) +

  # Draw bootstrap for Left-triangle
  geom_ribbon(data = ~get_boot_conf(filter(.x, 
                                           conv_or_not == "Visual", 
                                           a == "Left-triangle"),
                                    log_effect_size,
                                    reject,
                                    times = 1000) %>%
                mutate(conv_or_not = "Visual"),
              aes(x = log_effect_size, 
                  ymin = lower, 
                  ymax = upper),
              fill = rev(rcartocolor::carto_pal(3, "Vivid"))[3],
              alpha = 0.2) +

  # Draw bootstrap for Right-triangle
  geom_ribbon(data = ~get_boot_conf(filter(.x, 
                                           conv_or_not == "Visual", 
                                           a == "Right-triangle"),
                                    log_effect_size,
                                    reject,
                                    times = 1000) %>%
                mutate(conv_or_not = "Visual"),
              aes(x = log_effect_size, 
                  ymin = lower, 
                  ymax = upper),
              fill = rev(rcartocolor::carto_pal(3, "Vivid"))[2],
              alpha = 0.2) +

  # Theme
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(3, "Vivid")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

>>>>

```{r}
# Bootstrap the dataset
boot_y <- function(dat, y, times = 100) {
  map_df(1:times, function(i) slice_sample(dat, n = nrow(dat), replace = TRUE) %>% mutate(boot_id = i))
}
```

@@

  # Draw bootstrap visual test power curves
  stat_smooth(geom = "line", 
              data = ~boot_y(.x, reject, times = 500), 
              aes(log_effect_size, 
                  reject, 
                  group = boot_id, 
                  col = "Visual test"), 
              method = "glm", 
              method.args = list(family = binomial), 
              se = FALSE, 
              alpha = 0.03) +

@@

```{r polypower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for non-linear patterns (uniform fitted values only). Main plot shows the power curves estimated using logistic regression, with dots indicating human evaluations of lineups. Surrounding curves of the visual test show 500 samples of bootstrap estimation.  Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much non-linearity in the residuals? For the RESET test this is around log effect size 1.5, but for the visual test it is around 3.5."}

@@

```{r heterpower, fig.height = 3, fig.width = 5, fig.cap = "Comparison of power between different tests for heteroskedasticity patterns (uniform fitted values only). Main plot shows the power curves, with dots indicating human evaluations of lineups. Surrounding curves of the visual test show 500 samples of bootstrap estimation. Small row of plots shows typical residual plots corresponding to specific effect sizes, marked by dashed lines in main plot. Where would you draw the line of too much heteroskedasticity in the residuals? For the BP test this is around log effect size 1.5, but for the visual test it is around 3."}

@@

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. Although the power of visual test under the discrete uniform distribution is the lowest in the case of non-linearity patterns, it has a relatively great power this time. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Visual tests have low power on lineups with the normal distribution and the lognormal distribution. This suggests the use of transformation of fitted values in reading residual plots. In the case of conventional tests, the discrete uniform distribution consistently doing great, followed by the uniform distribution, the normal distribution and the lognormal distribution.

@@

```{r different-x-dist-poly-power, fig.cap = "Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. Each power curve of the uniform distribution is surrounded by a grey line to indicate it is the baseline in its group. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power for non-linearity patterns in residual plots. For heteroskedasticity patterns, normal and lognormal distributions produce lower power."}

@@

```{r poly-power-j, fig.height = 4, fig.cap = "Comparison of power of conventional tests and visual tests on lineups containing four different non-linearity shapes. "}

@@

@@

heter_conv_sim %>%
  filter(x_dist == "uniform") %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(reject = as.numeric(BP_p_value < 0.05)) %>%
  mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>%
  mutate(conv_or_not = "Conventional") %>%
  bind_rows(vi_survey %>%
  filter(!attention_check) %>%
  filter(!null_lineup) %>%
  filter(x_dist == "uniform", type == "heteroskedasticity") %>%
  mutate(log_effect_size = log(effect_size)) %>%
  mutate(reject = as.numeric(p_value < 0.05)) %>%
  mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>%
  mutate(conv_or_not = "Visual")) %>%
  mutate(a = factor(a, levels = c("Left-triangle", "Butterfly", "Right-triangle"))) %>%
  select(a, log_effect_size, reject, conv_or_not) %>%
  ggplot() +
  geom_smooth(aes(log_effect_size, reject, col = a), 
              se = FALSE, 
              method = "glm",
              method.args = list(family = binomial())) +
  xlab(expression(log[e] (Effect_size))) +
  ylab("Power") +
  scale_color_manual(values = rcartocolor::carto_pal(3, "Vivid")) +
  theme_light() +
  theme(legend.position = "bottom") +
  facet_wrap(~conv_or_not) +
  labs(col = "Heteroskedasticity shape")
```

### 2023-01-24 paper_comparison/paper_comparison/paper_comparison.tex
update color pal for the last section of the result
small changes

\caption{Comparison of power of conventional tests and visual tests on lineups with different fitted value distributions. Each power curve of the uniform distribution is surrounded by a grey line to indicate it is the baseline in its group. Conventional tests are more sensitive than visual tests in regards of the type of the simulated model and the fitted value distribution. For non-linearity patterns, visual tests show lower power with discrete uniform distribution, possibly due to the difficulty in reading discreteness. For heteroskedasticity patterns, because of the unevenness, visual tests have lower power with normal and lognormal distribution.}\label{fig:different-x-dist-poly-power}
\end{figure}

>>>>

\caption{Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. Each power curve of the uniform distribution is surrounded by a grey line to indicate it is the baseline in its group. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power for non-linearity patterns in residual plots. For heteroskedasticity patterns, normal and lognormal distributions produce lower power.}\label{fig:different-x-dist-poly-power}
\end{figure}
